---
title: "PA1_template"
author: "Pedro Porrino"
date: "8/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

# Assessment Course 5 Project 1

This script collects data from an activity monitoring device and explore data on
steps.
The sequence of analysis is:

## Loading the data

```{r readData}
activity <- read.csv("activity.csv")
```

## Mean total number of steps taken per day
To answer this question, data are grouped by date, the sum of steps per day
is computed and the information is plotted. Coordenates are flipped in order
to show tidier date labels. After that, mean and median are calculated. As NA values
have been ignored and not removed, for dates with intervals containing NAs, the sum
is NA and ggplot_col gives a warning saying 8 rows have been remove (meaning there are
8 days with NAs) These 8 days are shown as empty bars in the plot. 
Mean and Median throw NA as any computational operation involving NAs give NA.

```{r}
days <- group_by(activity, date) 
daily <- summarise(days, steps_day=sum(steps))

(ggplot(daily)+geom_col(mapping=aes(x=date, y=steps_day))+coord_flip())

(steps_day_mean <- mean(daily$steps_day))
(steps_day_median <- median(daily$steps_day))
```

## Average daily activity pattern
To see the daily activity pattern, a time series plot is tried to be drawn based on 5 min
intervals. For that, now grouping is done by intervals and mean of steps is
calculated for each interval

```{r}
intervals <- group_by(activity, interval)
intervaly <- summarise(intervals, avg=mean(steps))

(ggplot(intervaly)+geom_line(mapping=aes(x=interval, y=avg)))
```

What happened? We see a warning saying that 288 rows have been removed (actually
all the interval groups) meaning that as there are NA for all the intrevals, mean stat
will be NA, and therefore graph will be empty.

Obviously, a determination of which interval on average across all the days contains
the maximum number of steps will throw same empty result:

```{r}
(interval_max <- intervaly$interval[which.max(intervaly$avg)])
```

If we do the same but just we explicit na.rm=true when we try to calculate mean steps:

```{r}
intervaly <- summarise(intervals, avg=mean(steps, na.rm=TRUE))
```

Now, when we plot intervals serie, we can see pattern and calculate interval
in which maximum number of steps are taken:

```{r}
(ggplot(intervaly)+geom_line(mapping=aes(x=interval, y=avg)))

(interval_max <- intervaly$interval[which.max(intervaly$avg)])
```

## Imputing missing values
I assume that when there are no available data, the best way to minimize noise
is to consider 0 steps for that period. With code below we determine number of NA
(we saw before distributed in 8 different days) and we set them to 0
in a new data set named activity_narm:

```{r naDeal}
(na <- length(activity$steps[is.na(activity$steps)]))

activity_narm <- activity
activity_narm$steps[is.na(activity_narm$steps)]=0 
```

If same analysis as before is performed now regarding total number of steps taken
per day as well as total mean and median all across days:

```{r}
days_narm <- group_by(activity_narm, date) 
daily_narm <- summarise(days_narm, steps_day=sum(steps))

(ggplot(daily_narm)+geom_col(mapping=aes(x=date, y=steps_day))+coord_flip())

(steps_day_mean_narm <- mean(daily_narm$steps_day))
(steps_day_median_narm <- median(daily_narm$steps_day))
```

if we check how many days have 0 steps:

```{r}
(filter(daily_narm, daily_narm$steps_day == 0))
```
We see that there are 8 complete days with 0 steps, what means that all intervals
for these days had NAs. Just because of this, there is no difference in the plot 
(8 empty rows because of NAs and 8 empty rows beacuse of 0 steps)
 
## Differences in activity patterns between weekdays and weekends

We want to figure out if there is any change in activity patterns depending on whether it's weekday o weekend. We create a new variable to classify observations in weekday or weekend:

```{r}
activity_narm <- mutate(activity_narm, day=weekdays(as.Date(activity$date)))
for (i in 1:length(activity_narm$day)) {
        if (activity_narm$day[i] %in% c("lunes", "martes", "miércoles", "jueves", "viernes")) {
                activity_narm$day[i]="weekday"} 
        else {
                activity_narm$day[i]="weekend"}
        }
```

Data are grouped now by day and interval; mean steps are calculated by days and plotted
using facets to differentiate between weekdays and weekends:

```{r}
intervals <- group_by(activity_narm, day, interval)
intervaly <- summarise(intervals, avg=mean(steps))

(ggplot(intervaly)+geom_line(mapping=aes(x=interval, y=avg, color=day))+facet_wrap(~ day, nrow=2))
```


